---
title: "Weather Events - Damage and Casualties"
author: "Peter Olsen"
date: "November 19, 2015"
output: html_document
---

##Data Processing
```{r}
# clear out global environment, 
rm(list=ls())  

setwd("~/Documents/datasciencecoursera/ReproducibleResearch/PA2")
d <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))

#load libraries
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)

```
#### Start Cleaning the Data

From the Document https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf 
page 6 we see that data for all weather types was not collected until January 1996.
I believe this will give an inaccuate picture of what events cause the most problems because it 
will show Tornados and other events that were recooded earlier as being more prevelant than they
really are.
```{r}
# so I am going to limit dates to those before jan 1996
rdate <- as.POSIXct("1/1/1996", "%m/%d/%Y", tz="")
dt <- d[which(strptime(as.character(d$BGN_DATE), "%m/%d/%Y %H:%M:%S") > rdate),]
```
Lets also get rid of data that does not have any health or economic loss.
I will do this by dropping rows that don't have any property, crop, injury or fatality data.

```{r}
#subset on events with damage to persons or property
dsub <- subset(dt, dt$PROPDMG > 0 | dt$CROPDMG > 0 | 
                   dt$FATALITIES > 0 | dt$INJURIES > 0)
```

The data has Crop damage and Property damage and multipliers in it.  I am going to calculate a total 
property damage value.  This will be property damage times the the multiplier plus the same for crop damage- crop damage times the multiplier.  The multipliers need to be convert from a letter code to a numeric value.
```{r}

##First MAKE CHANGE TO Incorrect value pointed out by TA on Discussion board.
dsub[dsub$REFNUM == 605943,"PROPDMGEXP"] <- "M" 
 
dsub$PROPX[dsub$PROPDMGEXP %in% c("K","k") ] <- 1000
dsub$PROPX[dsub$PROPDMGEXP %in% c("M","m") ] <- 1000000
dsub$PROPX[dsub$PROPDMGEXP %in% c("B","b") ] <- 1000000000
# now calculate and add a complete property damage value
dsub <- mutate(dsub, PropertyDamage = PROPX * PROPDMG)  

#Do the same for crops
dsub$CROPX[dsub$CROPDMGEXP %in% c("K","k") ] <- 1000
dsub$CROPX[dsub$CROPDMGEXP %in% c("M","m") ] <- 1000000
dsub$CROPX[dsub$CROPDMGEXP %in% c("B","b") ] <- 1000000000
# now calculate and add a complete property damage value
dsub <- mutate(dsub, CropDamage = CROPX * CROPDMG) 

```

For my analysis I will be using a total for crop damages and property damages called EconCost.  
This will represent the total recorded cost of a weather event.
```{r}
dsub <- mutate(dsub, EconCost = CropDamage + PropertyDamage)
```

For my analysis I will sum the Fatalities and Injuries into a value called Casualties.  I will use this number to represent the recorded harm to human health caused by a weather event.

```{r}
dsub <- mutate(dsub, Casualties=  FATALITIES + INJURIES)
```
At this point I will clean out the columns of data I will no longer need for my analysis.

```{r}
data <- dsub[ , c(8,42, 43)]

c_names <- c("EventType","EconCost", "Casualties")
colnames(data) <- c_names

## remove uneeded data from memory
rm(dt)
rm(dsub)
rm(d)
```
####Event Type names
The event type names that are supposed to be used can be seen here:
https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf 
page 6
I have added them to a vector to use for verification purposes.
```{r}
#list of valid names from report
ValidNames <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", 
"Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil","Dust Storm","Excessive Heat", 
"Extreme Cold/Wind Chill", "Flash Flood", 
"Flood", "Frost/Freeze", "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow", 
"High Surf", "High Wind", "Hurricane (Typhoon)", "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", 
"Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", 
"Rip Current", "Seiche", "Sleet", "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", 
"Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", 
"Wildfire", "Winter Storm", "Winter Weather")
```


There are a lot of Event type names that do not meet what is stated:We can see some of them here.
```{r}
head(unique(data[,1]), 20)
```
The next batch of code will correct the Event Type Names so they will be consistent which will help us analys them.  I have commented the code for each step.

```{r}
## trim extra spaces
data[,1] <- str_trim(as.character(data[,1]), side = "both")

## update Thunderstorm before other changes
data[,1] <- sub("tstm","thunderstorm", data[,1], ignore.case = TRUE)

##GET Captializations right
data[,1] <- tolower(as.character(data[,1]))  ## start with all lower case
## functions to Uppercase first letter
capwords <- function(s, strict = FALSE) {
        cap <- function(s) paste(toupper(substring(s, 1, 1)),
{s <- substring(s, 2); if(strict) tolower(s) else s},
sep = "", collapse = " " )
sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
}
## Capitalize words with / before
capwordsSlash <- function(s, strict = FALSE) {
        cap <- function(s) paste(toupper(substring(s, 1, 1)),
{s <- substring(s, 2); if(strict) tolower(s) else s},
sep = "", collapse = "/" )
sapply(strsplit(s, split = "/"), cap, USE.NAMES = !is.null(names(s)))
}
# run functions
data[,1] <- sapply(data[,1], capwords)
data[,1] <- sapply(data[,1], capwordsSlash)

#Make some changes to Event names with sub
data[,1] <- sub("Thunderstorm Wind.+", "Thunderstorm Wind", data[,1], ignore.case = FALSE, perl = TRUE
    )
data[,1] <- sub(".*Snow.*", "Heavy Snow", data[,1], ignore.case = FALSE, perl = TRUE
)

data[,1] <- sub(".*wind.*", "High Wind", data[,1], ignore.case = TRUE, perl = TRUE
)
data[,1] <- sub(".*(hurricane|typhoon).*", "Hurricane (Typhoon)", data[,1], ignore.case = TRUE, perl = TRUE
)

data[,1] <- sub(".*Ice Jam Flood \\(minor.*", "Flood", data[,1], ignore.case = TRUE)

# use the event update table to make more changes to the Event types to get them to match Valid Event names.
updateTable <- read.csv("EventUpdate.csv", stringsAsFactors = FALSE)


for (j in 1:nrow(data)){
       if (data[j,][[1]] %in% ValidNames)
               next
    for (i in 1:nrow(updateTable)){
        if (grepl (updateTable[i,][[1]], data[j,][[1]]) )
                data[j,][[1]] <- updateTable[i,][[2]]        
    }
}
```

